\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2016-01-12 nnt_manu.tex
prescreening revision done

Training a neural network model requires identifying the optimal weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 80\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages('NeuralNetTools')
R> library(NeuralNetTools)
\end{Sinput}
\end{Schunk}

@@

A common approach for data preprocessing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  A sample dataset that follows this format is included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and five columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  Additional datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.  The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.

\begin{Schunk}
\begin{Sinput}
R> set.seed(123)
R> library(RSNNS)
R> x <- neuraldat[, c('X1', 'X2', 'X3')]
R> y <- neuraldat[, 'Y1']
R> mod1 <- mlp(x, y, size = 5)
R> library(neuralnet)
R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
R> library(nnet)
R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
\end{Sinput}
\end{Schunk}

@@

An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}. The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables. The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method. However, it is one of few methods used to evaluate sensitivity in neural networks.

The \code{lekprofile} function evaluates the effects of input variables by returning a plot of model predictions across the range of values for each variable.  The remaining explanatory variables are held constant when evaluating the effects of each input variable.  The \code{lekprofile} function provides two options for setting constant values of unevaluated explanatory variables.  The first option follows the original profile method by holding unevaluated variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}). This is implemented by creating a matrix where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables. All explanatory variables are held at a chosen constant value (e.g., median), while the variable of interest is sequenced from its minimum to maximum value across the range of observations. This matrix is then used to predict values of the response variable from a fitted model object. This is repeated for each explanatory variable to obtain all response curves.  Constant values are set in \code{lekprofile} by passing one or more values in the range 0--1 to the \code{group\_vals} argument.  The default holds variables at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles (i.e., \code{group\_vals = c(0, 0.2, 0.4, 0.6, 0.8, 1)}).

A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data. Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable). The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one. The centers (means) of the clusters are then used as constant values for the unevaluated variables.  \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.  Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.  Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).  For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.  Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
\begin{Schunk}
\begin{Sinput}
R> lekprofile(mod3)
R> lekprofile(mod3, group_show = TRUE)
R> lekprofile(mod3, group_vals = 6)
R> lekprofile(mod3, group_vals = 6, group_show = TRUE)
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\centering
\subfloat[][Response with quantiles]{
\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
\label{fig:plotlek_sens1}
}

@@

Several conclusions can be made from \cref{fig:flightimp}.  First, delays in arrival time are negatively related to distance and positively related to departure delays and travel air.  That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.  No conclusions can be made for the other variables because the bounds of uncertainty included zero.  Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).  This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.  Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.  This suggests that a single model can provide misleading information such that several models may be required to improve the certainty of conclusions.  Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
\begin{figure}[!h]
\centering
\subfloat[][Networks with one node]{
\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
\label{fig:flightimp1}
}

>>>>

Training a neural network model requires identifying the optimal weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citet{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a backpropagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  A fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a sum of products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 80\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class).  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("NeuralNetTools")
R> library("NeuralNetTools")
\end{Sinput}
\end{Schunk}

@@

A common approach for data pre-processing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  A sample dataset that follows this format is included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and five columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  Additional datasets can be pre-processed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.  The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.

\begin{Schunk}
\begin{Sinput}
R> set.seed(123)
R> library("RSNNS")
R> x <- neuraldat[, c('X1', 'X2', 'X3')]
R> y <- neuraldat[, 'Y1']
R> mod1 <- mlp(x, y, size = 5)
R> library("neuralnet")
R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
R> library("nnet")
R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
\end{Sinput}
\end{Schunk}

@@

An alternative approach to evaluate variable interactions in a neural network is the Lek profile method \citep{Lek96,Gevrey03}. The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables. The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method. However, it is one of few methods used to evaluate sensitivity in neural networks.

The \code{lekprofile} function evaluates the effects of input variables by returning a plot of model predictions across the range of values for each variable.  The remaining explanatory variables are held constant when evaluating the effects of each input variable.  The \code{lekprofile} function provides two options for setting constant values of unevaluated explanatory variables.  The first option follows the original profile method by holding unevaluated variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}). This is implemented by creating a matrix where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables. All explanatory variables are held constant (e.g., median) while the variable of interest is sequenced from its minimum to maximum. This matrix is then used to predict values of the response variable from a fitted model object. This is repeated for each explanatory variable to obtain all response curves.  Constant values are set in \code{lekprofile} by passing one or more values in the range 0--1 to the \code{group\_vals} argument.  The default holds variables at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles (i.e., \code{group\_vals = c(0, 0.2, 0.4, 0.6, 0.8, 1)}).

A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data. Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable). The second option holds unevaluated variables at means defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer greater than one. The centers (means) of the clusters are then used as constants for the unevaluated variables.  \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.  Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more interpretable response curves.  Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).  For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.  Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
\begin{Schunk}
\begin{Sinput}
R> lekprofile(mod3)
R> lekprofile(mod3, group_show = TRUE)
R> lekprofile(mod3, group_vals = 6)
R> lekprofile(mod3, group_vals = 6, group_show = TRUE)
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\centering
\subfloat[][Response with quantiles]{
\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
\label{fig:plotlek_sens1}
}

@@

Several conclusions can be made from \cref{fig:flightimp}.  First, delays in arrival time are negatively related to distance and positively related to departure delays and air time.  That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.  No conclusions can be made for the other variables because the bounds of uncertainty included zero.  Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).  This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.  Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.  This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.  Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
\begin{figure}[!h]
\centering
\subfloat[][Networks with one node]{
\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
\label{fig:flightimp1}
}
